---
title: "Advent of Code (Day 14 - 2021)"
author: "Francisco Yirá Albornoz Cambiaso"
date: "14-12-2021"
output: rmdformats::downcute
---

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(here)
library(testthat)
```


## Description of the problem
https://adventofcode.com/2021/day/14

## Load input
```{r}
template <- "OHFNNCKCVOBHSSHONBNF"
rules_str <- read_file(here("day14_input", "rules.txt"))
```


## Proof of concept
```{r}
# Create two vectors: 1
# 1) with strings to find
# 2) with strings to insert
vector_rules <- 
  str_split(rules_str, "\r\n") %>% 
  unlist()

vector_rules
```
```{r}
list_rules <- 
  vector_rules %>%
  str_split(" -> ")

list_rules[1:5]
```

```{r}
input_strings <- 
  list_rules %>% map(1) %>% unlist()

input_strings
```
```{r}
to_insert_strings <- 
  list_rules %>% map(2) %>% unlist()

to_insert_strings
```
```{r}
list_replacement <- 
  as.list(to_insert_strings) %>% 
  set_names(input_strings)

list_replacement[1:5]
```
Proof of concept:
```{r}
list_replacement["SV"]
```
Now iterate over the input (`template`) to apply the insertion rules:
```{r}
new_string <- str_sub(template, 1, 1)

for (i in seq(str_length(template)-1)) {
  sub_string <- template %>% str_sub(i, i+1)
  
  # Extract the string to insert from `list_replacement`
  # And skip if it doesn't exist
  string_to_insert <- list_replacement[[sub_string]]
  if (is.null(string_to_insert)) next
  
  new_string <- paste0(new_string,
                       string_to_insert,
                       str_sub(template, i+1, i+1))
}

new_string
```

It seems good as a proof of concept, but some validation is necessary.

## Testing
First, let's create functions to parametrize what I've done until now:
```{r create_list_pair_insertion}
create_list_pair_insertion <- function(filename_rules) {
  rules_str <- read_file(filename_rules)
  
  vector_rules <-
    str_split(rules_str, "\r\n") %>%
    unlist()
  
  list_rules <-
    vector_rules %>%
    str_split(" -> ")
  
  input_strings <-
    list_rules %>% map(1) %>% unlist()
  
  to_insert_strings <-
    list_rules %>% map(2) %>% unlist()
  
  list_replacement <-
    as.list(to_insert_strings) %>%
    set_names(input_strings)
  
  list_replacement
  
}
```

```{r pair_insertion}
pair_insertion <- function(template, list_pair_insertion) {
  new_string <- str_sub(template, 1, 1)
  
  for (i in seq(str_length(template) - 1)) {
    sub_string <- template %>% str_sub(i, i + 1)
    
    string_to_insert <- list_pair_insertion[[sub_string]]
    if (is.null(string_to_insert)) next
    
    new_string <- paste0(new_string,
                         string_to_insert,
                         str_sub(template, i + 1, i + 1))
  }
  
  new_string
}
```

Now a unit test using example posted in the instructions:
```{r}
test_that("Function pair_insertion produces correct answers",
          {
            list_example <- create_list_pair_insertion(here("day14_input",
                                                            "rules_example.txt"))
            
            insertion1 <- pair_insertion("NNCB", list_example)
            expect_equal(insertion1, "NCNBCHB")
            
            insertion2 <- pair_insertion(insertion1, list_example)
            expect_equal(insertion2, "NBCCNBBBCBHCB")
            
            insertion3 <- pair_insertion(insertion2, list_example)
            expect_equal(insertion3, "NBBBCNCCNBBNBNBBCHBHHBCHB")
            
            insertion4 <- pair_insertion(insertion3, list_example)
            expect_equal(insertion4, "NBBNBNBBCCNBCNCCNBBNBBNBBBNBBNBBCBHCBHHNHCBBCBHCB")
          })
```
## Solution Part 1
Now I can (confidently) repeat the insertion 10 times on the puzzle input:
```{r}
result <- template
for (i in 1:10) {
  result <- pair_insertion(result, list_replacement)
}
```

Now count frequency of elements:
```{r}
df_count_elements <- 
  tibble(LETTERS,
       count= str_count(result, LETTERS)) %>% 
  filter(count!=0) %>% 
  arrange(desc(count))

df_count_elements
```

Doing the substraction:
```{r}
counts <- df_count_elements[["count"]]

counts[1] - counts[length(counts)]
```
[ ] TODO: optimize the function in order to be performant enough to do 40 steps.

